---
title: "Tree Based Variable Selection"
author: "J. Sebastian Paez"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r message=FALSE, warning=FALSE, paged.print=FALSE}
require(Seurat)
require(ggplot2)
require(plyr)
require(dplyr)
require(ranger)
require(partykit)
require(caret)


data("pbmc_small")
```



```{r}

#' Gets a data frame from a seurat object
#' 
#' Provided a seurat object, returs a data frame of the count values, being 
#' the columns each 'gene' and the rows each UMI/cell.
#'
#' It returns only the genes annotated as variable and the identity column.
#'
#' @param seurat A seurat object
#'
#' @return
#' @export
#'
#' @examples
#' > as.data.frame.Seurat(pbmc_small)[1:3,1:3]
#'                    LTB EAF2 CD19
#' ATGCCAGAACGACT 6.062788    0    0
#' CATGGCCTGTGCAT 6.714813    0    0
#' GAACCTGATGAACC 7.143118    0    0
#' @importFrom Seurat FetchData GetIdent
as.data.frame.Seurat <- function(seurat) {
    tmp <- Seurat::FetchData(seurat, vars.all = seurat@var.genes)
    tmp <-  as.data.frame(tmp, stringsAsFactors = FALSE)

    tmp$ident <- Seurat::GetIdent(
      seurat, uniq = FALSE, cells.use = rownames(tmp))
    return(tmp)
}


pbmc_df <- as.data.frame.Seurat(pbmc_small)

# Some minor renaming is required for most methods, since some genes have - in 
# their names, and we require R-valid names.

original_names <- colnames(pbmc_df)
newnames <- make.names(original_names)
colnames(pbmc_df) <- newnames

```


## Most Basic Form of Decision Tree based variable selection


```{r}
tree <- ctree(ident ~ ., pbmc_df)
tree

varimp(tree)
plot(tree)
```

which could be plotted later by selecting the genes that worked best.
Also note how only imprtances are given to the genes that were used in the end.

```{r}
ggplot2::qplot(x = pbmc_df$PPBP, y = pbmc_df$GNLY, colour = pbmc_df$ident)
ggplot2::qplot(x = pbmc_df$PPBP, fill = pbmc_df$ident, geom = "density")
ggplot2::qplot(x = pbmc_df$GNLY, fill = pbmc_df$ident, geom = "density")

```


## Most Basic Form of Random forest based variable selection

```{r}

forest <- ranger::ranger(
  ident ~ ., data =  pbmc_df,
  importance = "impurity_corrected")

forest

df_importances <- as.data.frame(
  ranger::importance_pvalues(forest)[
      ranger::importance_pvalues(forest)[, "pvalue"] < 0.05, ])

head(df_importances[order(df_importances[, "importance"]), ] )


dim(df_importances)
```


## Adding_weights for under-represented clusters


```{r}

# This add weihts to the clusters in proportion to the inverse of their abundance
# TODO find a better way to calculate the weights ...
weights_vars <- ceiling(as.numeric(100 / table(pbmc_df$ident)))


forest <- ranger::ranger(
  ident ~ ., data =  pbmc_df,
  importance = "impurity_corrected",
  class.weights = weights_vars)

forest

df_importances <- as.data.frame(
  ranger::importance_pvalues(forest)[
      ranger::importance_pvalues(forest)[, "pvalue"] < 0.05, ])

head(df_importances[order(df_importances[, "importance"]), ] )

dim(df_importances)

```

Note that due to the nature of the random forest, it cannot be plotted ...
but importances can

```{r}
ggplo2::qplot(x = df_importances$importance, y = rownames(df_importances))
```

.... since factors are weird in R we need to reorder the levels for the plot to
make sense ...

```{r}
genes <- forcats::fct_reorder(
  rownames(df_importances),
  df_importances$importance)

ggplo2::qplot(x = df_importances$importance, y = genes)
```



## Packing all in a function

```{r}
find_cluster_classif_forest <- function(cluster_id,
                                               class_df,
                                               markers,
                                               report_n = 4) {
    print(paste0("Cluster: ", cluster_id))
    foo2 <- markers %>%
        unique() %>% {
          class_df[, colnames(class_df) %in% c(., "ident")]
          }

    in_cluster <- foo2$ident == cluster_id
    out_cluster <- !in_cluster

    num_outcluster <- sum(out_cluster)
    num_incluster <- sum(in_cluster)

    foo2$ident <- foo2$ident == cluster_id
    
    # This section alleviates slightly the problem when the cluster is extremely
    # small compared to the non-cluster points.
    # TODO  add a warning whenever this happens...
    if (100 < (num_outcluster / num_incluster)) {
      foo2 <- foo2[c(which(foo2$ident),
                     sample(which(!foo2$ident), num_incluster * 100)), ]
    }

    foo2$ident <- factor(make.names(foo2$ident))

    forest <- ranger::ranger(
        ident ~ .,
        data = foo2,
        importance = "impurity_corrected",
        num.trees = 2000)

    df_importances <- as.data.frame(
      ranger::importance_pvalues(forest))

    df_importances$marker <- rownames(df_importances)

    markers <- dplyr::top_n(df_importances, report_n, `importance`)$marker

    return(
      list(markers = markers,
           tree = forest,
           cluster = cluster_id,
           df_importances = df_importances))
}


tmp <- find_cluster_classif_forest(
    cluster_id = "1",
    class_df = pbmc_df,
    markers = pbmc_small@var.genes,
    report_n = 5 )

lapply(tmp, summary)

```

## Now applying that function to all the clusters

```{r fig.height=3, fig.width=3}

classif_info <- lapply(unique(pbmc_df$ident),
       function(x) find_cluster_classif_forest(
         cluster_id = x,
         class_df = pbmc_df,
         markers = rownames(df_importances)))


# This name conversion step carried out by `plyr::mapvalues`
# is required because `FeaturePlot` uses the
# original seurat object which still has the old names
for (i in classif_info) {
   i$markers %>%
    plyr::mapvalues(
      from = newnames,
      to = original_names,
      warn_missing = FALSE) %>% {
       for (marker in .) {
            FeaturePlot(
                object = pbmc_small,
                marker,
                cols.use = viridis::viridis(7),
                reduction.use = "tsne")
        }
   }
}
```


# Adding cross validation ...

Note how this is A LOT more computationlly expensive than running single models


```{r}

tc <- caret::trainControl(method = "CV", number = 5)

models <- caret::train(
    ident ~ .,
    data = pbmc_df,
    metod = "ranger",
    trControl = tc)

models

caret::varImp(models)


## Now for a glmnet
tc <- caret::trainControl(method = "CV", number = 5)

models <- caret::train(
  ident ~ .,
  method = "glmnet",
  data = pbmc_df,
  trControl = tc)

models

caret::varImp(models)
```


